---
title: "An Ode to Waffle Fries"
author: William E. Fondrie
date: '2019-07-21'
slug: an-ode-to-waffle-fries
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE, 
                      echo = FALSE,
                      dpi = 100,
                      dev.args = list(bg="transparent"))
```

```{r load_libraries}
suppressMessages(library(tidyverse, quietly = TRUE))
source("../../../src/theme_blog.R")
theme_set(theme_blog())
my_pal = c(theme_colors$fg, theme_colors$hl)
#set.seed(1251)
set.seed(1)
```

My family and I love to eat at Chick-fil-A. While we will forever miss the 
Chick-fil-A Cow Calendar (which was sadly ended in 2018) we have enjoyed their 
new rewards program. This short post is entirely about getting the best value 
from it.

## The rewards program  
The [Chick-fil-A One](https://www.chick-fil-a.com/one) rewards program is pretty
simple in how it works: 
You receive points for each dollar spent at Chick-fil-A, which can then
be redeemed for a variety of menu items at a latter date. However, I noticed
that the number of points required for a few items was not proportional to 
the item's menu price.  

**So, what is the most efficient way to spend my points?**

Naturally, I had to know.


## Point costs are roughly proportional to dollar values

To answer this question, I manually looked up the number of points required 
for each menu item and their corresponding menu price at our local Chick-fil-A. 
Unsurprisingly, it seems that the point cost of an item is
largely proportional to the item's dollar value. There are some interesting
deviations from this though: The three salads require disproportionately more
points to redeem than the other items.

```{r cfa_data}
cfa <- tibble(item = c("Cookie",
                       "Hash Browns",
                       "Icedream Cone",
                       "Waffle Fries",
                       "Fountain Drink",
                       "Sweet Tea",
                       "Unsweetend Tea",
                       "Chicken Biscuit",
                       "Diet Lemonade",
                       "Lemonade",
                       "Fruit Cup",
                       "4 ct Chick-n-Minis",
                       "8 ct Nuggets",
                       "Chicken Sandwich",
                       "Egg White Grill",
                       "Hash Brown Scramble Bowl",
                       "Hash Brown Scramble Burrito",
                       "Small Frosted Beverage",
                       "Small Milkshake",
                       "Spicy Chicken Sandwich",
                       "Deluxe Chicken Sandwich",
                       "Large Frosted Beverage",
                       "Large Milkshake",
                       "8 ct Grilled Nuggets",
                       "Grilled Chicken Sandwich",
                       "Spicy Deluxe Chicken Sandwich",
                       "4 ct Chic-n-Strips",
                       "Grilled Cool Wrap",
                       "Grilled Chicken Club",
                       "Cobb Salad",
                       "Market Salad",
                       "Spicy Southwest Salad"),
              price = c(1.49, 1.49, 1.65, 2.29, 2.15, 2.15, 2.15, 3.35, 2.45,
                        2.45, 3.85, 4.39, 4.59, 4.49, 4.69, 4.69, 4.69, 4.09,
                        4.09, 4.85, 5.09, 4.65, 4.65, 5.49, 6.19, 5.45, 6.29, 
                        7.59, 7.89, 9.39, 9.39, 9.39),
              points = c(150, 150, 150, 200, 250, 250, 250, 350, 350, 350,
                         450, 500, 500, 500, 550, 550, 550, 550, 550, 550,
                         650, 650, 650, 700, 750, 750, 800, 900, 950, 1500,
                         1500, 1500))

cfa %>% 
    ggplot(aes(x = price, y = points)) +
    geom_abline(slope=100, size = 1, color = theme_colors$hl[1]) +
    geom_point(color = theme_colors$fg) +
    labs(x = "Menu Price (USD)",
         y = "Point Cost",
         title = "Chick-fil-A Reward Items",
         subtitle = "Based on prices from Seattle's Aurora Ave store") +
    annotate("text", x = 9.39, y = 1500, hjust = 1.2,
             vjust = 0.5, label = "Salads", color = theme_colors$fg) +
    annotate("text", x = 6, y = 600, hjust = -0.2,  vjust = 1, 
             label = "100 points / dollar", color = theme_colors$hl[1]) +
    annotate("text", 2.29, y = 200, hjust =-0.1, vjust = 1, 
             label = "Waffle Fries", color = theme_colors$fg)
```

## Waffle fries are king  

One thing we can do is look at which items give your points the most purchasing
power. In this case, the medium waffle fries are the clear winner coming in at 
a value of $0.01145 per point redeemed. 

```{r items, fig.height=7}
cfa %>%
    mutate("point value" = price/points * 100,
           item = fct_reorder(item, `point value`)) %>% 
    ggplot(aes(x = item, y = `point value`)) +
    geom_point(size = 2, color = theme_colors$fg, size = 2) + 
    geom_segment(aes(x = item, xend = item, y = 0.6, yend = `point value`),
                 color = theme_colors$fg, size = 1) +
    theme(axis.title.y = element_blank()) +
    labs(y = "Point Value (cents)",
         title = "Point Values of Reward Items",
         subtitle = "Based on prices from Seattle's Aurora Ave Store.") +
    scale_y_continuous(limits = c(0.6, 1.2), expand = c(0, 0)) +
    coord_flip()
```

## Conclusion

Since I always get a spicy chicken sandwich and fries anyway, I'll now
redeem points for the fries instead of the sandwich. Unfortunately, your mileage
with this analysis may vary---the results may be different for other locations
and if discounts from combos were considered.

It may not seem like much, but if you're like me, this helps the postdoc dollars
go a little further.



